---
title: Analyse questionnaire and surveys in R
author: Masumbuko Semba
date: '2021-11-25'
slug: []
categories:
  - ggplot2
tags:
  - survey
  - questionnaire
  - interactive
  - marine
  - Masumbuko Semba
description: ''
thumbnail: ''
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")

```

## Introduction
This post offers some technique on how to analyse data from a surveys and questionnaires in R, provides tips on visualizing survey data, and exemplifies how survey and questionnaire data can be analyzed. 

Questionnaires and surveys are widely used in research and thus one of the most common research designs. 
Questionnaires elicit three types of data:

+ Factual
+ Behavioral
+ Attitudinal

While factual and behavioral questions are about what the respondent is and does, attitudinal questions tap into what the respondent thinks or feels.

The advantages of surveys are that they * offer a relative cheap, quick, and effective way to collect (targeted) data from a comparatively large set of people; and * that they can be distributed or carried out in various formats (face-to-face, by telephone, by computer or via social media, or by postal service).

Disadvantages of questionnaires are that they are prone to providing unreliable or unnatural data. Data gathered via surveys can be unreliable due to the social desirability bias which is the tendency of respondents to answer questions in a manner that will be viewed favorably by others. Thus, the data that surveys provide may not necessarily be representative of actual natural behavior.

In this section, we will discuss what needs to kept in mind when designing questionnaires and surveys, what pieces of software or platforms one can use, options for visualizing questionnaire and survey data, statistical methods that are used to evaluate questionnaire and survey data (reliability), and which statistical methods are used in analyzing the data.

## packages
s
```{r}
library(tidyverse)
library(likert)
```

## load the dataset


```{r, echo=FALSE}
qns =  read_csv("d:/semba/BIOSTATISTIC-Responses.csv")

```

```{r, eval=FALSE}
qns =  read_csv("BIOSTATISTIC-Responses.csv")

```

```{r}
qns %>% 
  glimpse()
```

## Visualizing survey data

Just as the data that is provided by surveys and questionnaires can take various forms, there are numerous ways to display survey data. In the following, we will have a look at some of the most common or useful ways in which survey and questionnaire data can be visualized. However, before we can begin, we need to set up our R session as shown below.

## Line graphs for Likert-scaled data

A special case of line graphs is used when dealing with Likert-scaled variables (we will talk about Likert scales in more detail below). In such cases, the line graph displays the density of cumulative frequencies of responses. The difference between the cumulative frequencies of responses displays differences in preferences. We will only focus on how to create such graphs using the ggplot environment here as it has an in-build function (ecdf) which is designed to handle such data.

The questionnaire dataset contains Likert-scaled variables about their competency in using computer. The response to the Likert item is numeric so that very incompetent would get the lowest (1) and high comppetent the highest numeric value (5).


```{r}
pc = qns %>% 
  # glimpse() %>% 
  select(degree = 9, competance = 10)

pc
```

The `pc` dataset set has only two columns: a column labeled degree which has three degree programs (German, Japanese, and Chinese) and a column labeled competence which contains values from 1 to 5 which represent values ranging from very incompetent to very competent. Now that we have data resembling a Likert-scaled item from a questionnaire, we will display the data in a cumulative line graph.

```{r}
pc %>% 
  ggplot(aes(x = competance, color = degree))+
  geom_step(aes(y = ..y..), stat = "ecdf")+
  scale_y_continuous(name = "Cumulative Density")+ 
  scale_x_discrete(limits = c("1","2","3","4","5"), 
                   breaks = c(1,2,3,4,5),
                   labels=c("Very incompetent", "Incompetent", 
                            "Neutral", "Competent", "Very competent")) + 
  ggsci::scale_color_d3()+
  theme_bw() +
  theme(legend.position = c(.12,.85))
```

<!-- The satisfaction of the German course was the lowest as the red line shows the highest density (frequency of responses) of very dissatisfied and dissatisfied ratings. The students in our fictitious data set were most satisfied with the Chinese course as the blue line is the lowest for very dissatisfied and “dissatisfied” ratings while the difference between the courses shrinks for “satisfied” and very satisfied. The Japanese language course is in-between the German and the Chinese course. -->

## Pie charts
Most commonly, the data for visualization comes from tables of absolute frequencies associated with a categorical or nominal variable. The default way to visualize such frequency tables are pie charts and bar plots. In a first step, we modify the data to get counts and percentages.

```{r}

degree = qns %>% 
  rename(degree = 9) %>% 
  group_by(degree) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count/sum(count)*100, 
         across(is.numeric, round, 1)) %>% 
  ungroup() %>% 
  mutate(degree = if_else(is.na(degree), "Unspecified", degree))
  

degree
```

```{r}
degree %>% 
  ggplot(aes(x = "", y = percent, fill = degree))+
  geom_col(color = "white")+
  coord_polar(theta = "y", start = 0)+
  theme_void()+
  ggsci::scale_fill_d3()
```

If the slices of the pie chart are not labelled, it is difficult to see which slices are smaller or bigger compared to other slices. This problem can easily be avoided when using a bar plot instead. This issue can be avoided by adding labels to pie charts. The labeling of pie charts is, however, somewhat tedious as the positioning is tricky. Below is an example for adding labels without specification.


```{r}
degree %>% 
  ggplot(aes(x = "", y = percent, fill = degree))+
  geom_col(color = "white")+
  coord_polar(theta = "y", start = 0)+
  geom_text(aes(y = percent, label = percent), color = "white")+
  theme_void()+
  ggsci::scale_fill_d3()
```

To place the labels where they make sense, we will add another variable to the data called “Position”.

```{r}
pdegree = degree %>% 
  arrange(desc(degree)) %>% 
  mutate(position = cumsum(percent) - 0.5*percent)

pdegree
```

Now that we have specified the position, we can include it into the pie chart.


```{r}
pdegree %>% 
  ggplot(aes(x = "", y = percent, fill = degree))+
  geom_col(color = "white")+
  coord_polar(theta = "y", start = 0)+
  geom_text(aes(y = position, label = percent), color = "white")+
  theme_void()+
  ggsci::scale_fill_d3()
```

We will now create separate pie charts for each gender. In a first step, we create a data set that does not only contain the degree levels and their frequency but also the gender.


```{r}
degree.gender = qns %>% 
  rename(degree = 9, gender = Gender) %>% 
  group_by(gender, degree) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count/sum(count)*100, across(is.numeric, round, 1)) %>% 
  arrange(desc(degree)) %>% 
  # ungroup() %>% 
  mutate(position = cumsum(percent) - 0.5*percent)

degree.gender
```

Let’s briefly inspect the new data set. Now that we have created the dataset, we can plot separate pie charts for each gender.

```{r}

degree.gender %>% 
  ggplot(aes(x = "", y = percent, fill = degree))+
  geom_col(color = "white")+
  geom_text(aes(y = position, label = percent), color = "white")+
  facet_wrap(~gender)+
  coord_polar(theta = "y", start = 0)+
  theme_void()+
  ggsci::scale_fill_d3()
    
```

## Bar plots
Like pie charts, bar plot display frequency information across categorical variable levels.

```{r}
degree.count = qns %>% 
  select(degree = 9, gender = Gender) %>% 
  group_by(degree) %>% 
  count()

degree.count
```


```{r}
degree.count %>% 
  ggplot(aes(x = degree, y = n, fill = degree))+
  geom_col()+
  geom_text(aes(x = degree, y = n-3, label = n), color = "white")+
  scale_y_continuous(name = "Frequency", expand = c(0, NA))+
  ggsci::scale_fill_d3()
```

Compared with the pie chart, it is much easier to grasp the relative size and order of the percentage values which shows that pie charts are unfit to show relationships between elements in a graph and, as a general rule of thumb, should be avoided.

Bar plots can be grouped which adds another layer of information that is particularly useful when dealing with frequency counts across multiple categorical variables. But before we can create grouped bar plots, we need to create an appropriate data set.


```{r}
degree.gender.count = qns %>% 
  select(degree = 9, gender = Gender) %>% 
  group_by(degree,gender) %>% 
  count()

degree.gender.count
```

We have now added Course as an additional categorical variable and will include gender as the “fill” argument in our bar plot. To group the bars, we use the command `position=position_dodge()`.

```{r}
degree.gender.count %>% 
  ggplot(aes(x = degree, y = n, fill = gender))+
  geom_col(position = position_dodge(.9))+
  geom_text(aes(x = degree, y = n-1, label = n), color = "white", position = position_dodge(.9))+
  scale_y_continuous(name = "Frequency", expand = c(0, NA))+
  ggsci::scale_fill_d3()+
  theme(legend.position = c(.085,.85))
```

```{r}


qns %>% 
  select(degree = 9, gender = Gender) %>% 
  ggplot(aes(x = degree))+
  geom_bar(fill = "maroon")+
  facet_wrap(~gender)

```


## multiple response

```{r}
likert.re = qns %>% 
  mutate(across(is.character, as.factor)) %>% 
  as.data.frame() %>%   
  dplyr::select(12,21,22) %>% 
  likert() 
  
  likert.re
```


```{r}
likert.re %>% 
  plot(ordered = F, wrap= 60)+
  ggsci::scale_fill_d3()

```

