---
title: Access Open Street Map features programmatically with osmdata package in R
author: Masumbuko Semba
date: '2021-11-13'
slug: []
categories:
  - ggplot2
  - R
tags:
  - simple feature
  - semba
  - Masumbuko Semba
description: ''
thumbnail: ''
bibliography: [../blog.bib]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE, comment = "")

load("d:/semba/theHinge/content/post/data/osm.RData")


```

**OpenStreetMaps** is a great source of *spatial* data. Most common programming languages have packages for downloading data from OSM. In this tutorial we are going to see how to download hosptial features data using R’s **osmdata** [@osmdata] package and plot it using **ggplot** [@ggplot], and interactively using **tmap** [@tmap]. This requires some knowledge of spatial data structures. We load the package we are going to use in our session.

```{r, eval=TRUE}
library(osmdata)
library(tidyverse)
library(patchwork)
library(sf)
library(tmap)
library(ggmap)
require(tmap)

```


## List recognized features in OSM

The osmdata package has a ton of awesome features. It appears that only a fraction of them are available for most cities, and that seems to be primarily those that are related to streets. We’ll focus on those today, but I hope they continue extending this package. The `available_features()` function list the features that are in the OSM.

```{r, eval=TRUE}
available_features()

```

Once we know the feature, we use  `avilable_tags` function to obtain all available features of the a particular class feature. FOr example, we want to know the different types of building that are found in 


```{r, eval=TRUE}
available_tags(feature = "building")
```

## Getting the Data

The first step requires downloading data from the OSM api using osmdata. First, I will need to set a bounding box for OSM to search for data with `getbb` function. Next,  download hosptial in Dar from a building class type with `add_osm_feature` function and finally convert the osm dataset into feature class with `osmdata_sf` function. The chunk below highlight the key step

```{r}

## hospital
dar.hosp = getbb(place_name = "dar es salaam") %>%
  opq()%>%
  add_osm_feature(key = "building", value = "hospital") %>%
  osmdata_sf()

 hosp = dar.hosp$osm_points 
 hosp
```

We then plot the hospital in Dar with tmap, which allows the user to interact with the map

```{r fig1, eval=TRUE, fig.cap="Interactive hosptials locations"}


tmap_mode(mode = "view")

tm_shape(shp = hosp)+
  tm_bubbles(size = .2)
```



sometimes you may need a basemap, we can obtain basemap from osmdata as the chunk highlight using a `get_map` function from **ggplot2** [@ggplot].

```{r}

## basemap
mad_map <- getbb(place_name = "dar es salaam") %>% 
  get_map(maptype="roadmap",source="osm",color="bw")


```


Then we can now use ggmap function of **ggmap** [@ggmap] to plot a basemap and overlay hospital with `geom_sf` function of **ggplot2** package and draw a map that show the distribution of hospital in Dar (Figure \@ref(fig:hosp2)). 

```{r hosp2, fig.cap="Distribution of hospitals in Dar es Salaam", eval=TRUE}

ggmap(ggmap = mad_map)+
  geom_sf(data = hosp,
          inherit.aes = FALSE)+
  theme(axis.title = element_blank())
  
```

## Listed literature

