---
title: Familiarize with date and time of Argo Floats data with lubridate package
author: Masumbuko Semba
date: '2020-07-05'
slug: []
categories:
  - R
tags:
  - lubridate
  - programming
  - semba
  - tidyverse
  - R
description: ''
thumbnail: ''
---



```{r , include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "")

require(tidyverse)
```


In this post we will learn to work with date and time data in R. We will use the **lubridate** package developed by Garrett Grolemund and Hadley Wickham ~@lubridate. This package makes it easy to work with dates and time. Let's us load the packages that we will use

```{r}
require(lubridate)
require(tidyverse)
require(magrittr)
require(oce)
```

## Data

```{r, echo=FALSE}
## to reduce the processing time, the loop in the chunk below was processed and the data saved in the working directory as .csv. This is loaded here but the chunk is hidden from view

argo.ctd.indian = vroom::vroom("../data/argo_indian.csv")

## compute the duration of floats from date variable
floats.duration = argo.ctd.indian %>% 
  filter(pressure == 5)  %>% 
  group_by(ID) %>% 
  summarise(start = first(time), 
            end = last(time), 
            period = interval(start, end) %>% as.duration() %>% as.numeric("years"), 
            count = n()) %>% arrange(count %>% desc())

```



We will use the profiles data from Argo within the Indian Ocean. The data was downloaded from the Coriolis Global Data Assembly Center site (ftp://ftp.ifremer.fr/ifremer/argo/) as NetCDF files. 

## Data processing

The argo profiles were converted from `.nc` format to data frame. The chunk below briefly describe each step. If you get stuck on the step, consult chapter \@ref() that describe looping in details. 

```{r, eval=FALSE, echo=FALSE}
# list the netcdf file names in the working directory

argo.file = dir("E:/Doctoral/udsm/Processing//argo_profile/",  #relative path of the directory
                recursive = TRUE, # recurse into directory
                pattern = "_prof.nc", # choose the names match this condition
                full.names = TRUE) # get name with its relative path

# preallocate a file to store processed ctd data
argo.ctd = NULL

# the first section of the loop run with i through the netcdf files
    for (i in 1:length(argo.file)){
        # read the files in sequence
      argo = read.argo(argo.file[i])%>%handleFlags()%>%argoGrid(p = seq(0,1000,5))
        # convert the argo list data into section
      argo.section = argo%>%as.section()
        # convert argo.section into list
      argo.list = argo.section[["station"]]

# the second section of the loop run with j through each argo list created by i
    for (j in 1:length(argo.list)){
      
    profile = argo.list[[j]]@data%>% # get each station profile data
      as.data.frame()%>%          # convert the data into daa frame
            # add variable of argo id, note this use i and not j
      mutate(argo = argo.file[i], 
              #add station variable
             station = argo.list[[j]][["station"]],
              #add date of sampling variable
             time = argo.list[[j]][["startTime"]]%>%as.Date(),
              #add longitude of station variable
             longitude = argo.list[[j]][["longitude"]], 
              #add latitude of station variable
             latitude = argo.list[[j]][["latitude"]], 
              #compute the salinity variable in each station
             density = argo.list[[j]]%>%swRho(eos = "gsw"))%>%
        # convert the data frame into tibble
      as.tibble()%>%
      # drop other variable of no interest
      select(argo, station, time, longitude, latitude,scan, pressure, 
             salinity = salinityAdjusted, temperature = temperatureAdjusted)
    #
    # separate the wmoid into into diferent variables
    profile = profile%>%separate(argo, c(1:7,"wmoid",8), sep = "/")%>%
      # keep the wmoid variable and drop the rest
      select(-c(1:8,10)) %>% rename(ID = 1)
    #
    #bind the argo.ctd with the data by rows
    argo.ctd = argo.ctd%>%bind_rows(profile)
  
    }
  }

```


There are 52 argo floats measured profiles of temperature and salinity as function of density between `r argo.ctd.indian %$% min(time)` and `r argo.ctd.indian %$% min(time)` and made a total of `r floats.duration %$% sum(count)` individual profiles.

Say you want create a column that show the durationof  each argo floats in the Indian Ocean. This information is important because it can help identify for on average how long does each float last or identify floats with the shortest or longest operation in the ocean. 

To accomplish this task and being able to answer those question, First, the argo floats were aggregated by id. Second, create two variable based on the Id, one variable contain the begin time of the float and the second variable is the end time of the variable. Third, compute the time interval and duration of the float based on the begin and end time. The table \@ref(tab:tab0899) show the sample of output resulted from the computation in the chunk below;

```{r, eval=FALSE}
floats.duration = argo.ctd.indian %>% 
  filter(pressure == 5)  %>% 
  group_by(ID) %>% 
  summarise(start = first(time), 
            end = last(time), 
            period = interval(start, end) %>% as.duration() %>% as.numeric("years"), 
            count = n()) %>% arrange(count %>% desc())
```

```{r tab0899}
floats.duration %>% 
  slice(1,seq(3,52, 6),52) %>%
  kableExtra::kable(format = "html", digits = 2, align = "c", 
                    caption = "The period and number of profiles made of randomly selected Argo floats",
                    col.names = c("Float ID", "Begin", "End", "Duration (years)", "Profile")) %>%
  kableExtra::column_spec(column = 1:3, width = "3cm") %>%
  kableExtra::column_spec(column = 4, width = "4cm") %>%
  kableExtra::add_header_above(c("", "Time of Argo Float" = 2, "", ""))
```

```{r}
floats.duration %>% filter(period < 2)
```


```{r}
argo2900564 = argo.ctd.indian %>% filter(ID == 2900564)

ggplot(data = argo2900564 %>% filter(pressure == 5), 
       aes(x = longitude, y = latitude, group = ID)) + geom_path()+geom_point() +
  metR::scale_y_latitude(ticks = 2, expand = c(.1,.1)) +
  metR::scale_x_longitude(ticks = 5, expand = c(.1,.1))
```


